---
title: "R Notebook"
output: html_notebook
---
##Roy Bengtsson Hangård
##Title:Datawrangling, module01
##Created:2024-09-30
##Last modified: 2024-10-03

* Loading the tidyverse package.

* Loading the dataset intoa dataframe

* Looking at the column specifications.

* Looking at the dataframe.
```{r}
library(tidyverse)
df_mothcount <- read_csv("MothCount.csv")
spec(df_mothcount)
df_mothcount
```
* Renaming the 3 variable names
* Creating a new dataframe
* Combining 2 columns into 2 new columns
* Col(morph) shows the observations morphology
* Col(count) sum up every individual for every sex-morph combination
* Looking at the new dataframe
```{r}
df_mothcount <- rename(df_mothcount,
                       sex = ...1,
                       dark = `Dark morph`,
                       light = `Light morph`)
df_mothcount_tidy <- pivot_longer(data = df_mothcount, 
                                  cols = dark:light,
                                  names_to = 'morph',
                                  values_to = 'count')
df_mothcount_tidy
```
* Loading a dataset into a dataframe
* Looking at dataframe and col.specifications
* Printing the entire dataframe
* Renaming col.(1) to 'ID'
* Combining the cols(pre.1:pre.5)
* Creating cols(period, measurements) from (pre.1:pre.5)
```{r}
df_OBK <- read_csv('OBrienKaiser.csv')
df_OBK
spec(df_OBK)
print.data.frame(df_OBK)

df_OBK %>% 
  rename(ID = ...1) %>%
  pivot_longer(cols = pre.1:fup.5, 
               names_to = "period",
               values_to = "measurements")
```
* Renaming col.(1) to 'ID'
* Combining the cols(pre.1:pre.5)
* Creating cols(period, measurements) from (pre.1:pre.5)
* Printing the entire dataframe
```{r}
df_OBK %>%
  rename(ID = ...1) %>%
  pivot_longer(cols = pre.1:fup.5,
               names_to = "period",
               values_to = "measurements") %>%
  print.data.frame(.) 
```
* Renaming col.(1) to 'ID'
* Combining the cols(pre.1:pre.5)
* Creating cols(period, measurements) from (pre.1:pre.5)
* Including only the last 10 observations
```{r}
df_OBK %>%
  rename(ID = ...1) %>%
  pivot_longer(cols = pre.1:fup.5,
               names_to = "period",
               values_to = "measurements") %>%
  tail(., n=10)
```
* Renaming col.(1) to 'ID'
* Combining the cols(pre.1:pre.5)
* Creating cols(period, measurements) from (pre.1:pre.5)
* Seperating col(period)
* Creating cols(period, time)
```{r}
df_OBK %>%
  rename(ID = ...1) %>%
  pivot_longer(cols = pre.1:fup.5,
               names_to = "period",
               values_to = "measurements") %>%
  separate(col = period,
           into = c("period", "time"),
           sep = "\\.")
```
* Renaming col.(1) to 'ID'
* Combining the cols(pre.1:pre.5)
* Creating cols(period, measurements) from (pre.1:pre.5)
* Seperating col(period)
* Creating cols(period, time)
* Converting col(time) character type to numeric type
```{r}
df_OBK %>%
  rename(ID = ...1) %>%
  pivot_longer(cols = pre.1:fup.5,
               names_to = "period",
               values_to = "measurements") %>%
  separate(col = period,
           into = c("period", "time")) %>%
  mutate(time = as.numeric(time))
```
* Creating a datframe
* Renaming col.(1) to 'ID'
* Combining the cols(pre.1:pre.5)
* Creating cols(period, measurements) from (pre.1:pre.5)
* Seperating col(period)
* Creating cols(period, time)
* Converting col(time) character type to numeric type
```{r}
df_OBK_tidy <- df_OBK %>%
  rename(ID = ...1) %>%
  pivot_longer(cols = pre.1:fup.5,
               names_to = "period",
               values_to = "measurements") %>%
  separate(col = period,
           into = c("period", "time")) %>%
  mutate(time = as.numeric(time))
```
* Loading a dataset to a dataframe
* Looking at col.specification
* Looking at dataframe
```{r}
tom_cities_df <- read_csv('Tom_cities_long.csv')
spec(tom_cities_df)
tom_cities_df
```
* Loading a dataset to a dataframe
* Looking at col.specification
* Looking at dataframe
```{r}
TOM_CITIES_Scotland_df2 <- read_csv('TOM_CITIES.csv')
spec(TOM_CITIES_Scotland_df)
TOM_CITIES_Scotland_df2
```
* Creating a dataframe
* Combining 10 columns
* Creating 2 new columns
```{r}
TOM_CITIES_Scotland_df3 <- TOM_CITIES_Scotland_df2%>%
  pivot_longer(cols = 'Day 1':'Day 10',
               names_to = "DAY",
               values_to = "Temperature")
```
* Creating a dataframe
* Combining 10 columns
* Creating 2 new columns
* Looking at dataframe
* Converting col(Day)character type to numeric type
* Dropping col(redundant) from the new dataframe
```{r}
NEW_CITIES_Scotland_df4 <-  TOM_CITIES_Scotland_df3%>% separate(col = DAY, into = c("redundant", "Day"))
NEW_CITIES_Scotland_df4
NEW_CITIES_Scotland_df4 %>% mutate(Day = as.numeric(Day))

NEW_CITIES_Scotland_df4 %>% select(-redundant) %>% 
  mutate(Day = as.numeric(Day))
```
* Creating a new dataframe
* Dropping col(redundant) from the new dataframe
* Creating a new dataframe
* Converting col(Day)character type to numeric type
* Looking at the dataframe
```{r}
TEMPS_CITIES_df <- NEW_CITIES_Scotland_df %>% 
  select(-redundant)
Cities_temps_df <- TEMPS_CITIES_df %>% 
  mutate(Day = as.numeric(Day))
Cities_temps_df
```
* Arranging variables in col(Temperature) in a descending order 2 times
* Kepping the dataframe's columns
```{r}
Cities_temps_df %>%
  arrange(desc(Temperature))

Cities_temps_df %>%
  arrange(desc(Temperature)) %>%
  distinct(City, .keep_all = TRUE)
```
* Filtering out 'Tuscaloosa from col(City)
* Arranging variables in col(Temperature) in a descending order
```{r}
Cities_temps_df %>%
  filter(City != "Tuscaloosa") %>%
  arrange(desc(Temperature))
```
* Filtering out 'Tuscaloosa from col(City)
* Arranging variables in col(Temperature) in an ascending order
```{r}
Cities_temps_df %>%
  filter(City != "Tuscaloosa") %>%
  arrange(Temperature)
```
* Filtering col(City) for 3 cities
* Arranging variables in col(Temperature) in an ascending order
```{r}
Cities_temps_df %>%
  filter(City %in% c("Cambridge","Falmouth","London")) %>%
  arrange(Temperature)
```
* Grouping the cities together under col(City)
* Calculating the temperature mean for each city
```{r}
  Cities_temps_df %>%
  group_by(City) %>%
  summarise(Temp_mean = mean(Temperature))
```
* Grouping the cities together under col(City)
* Calculating temperature mean and standard deviation from col(Temperature)
```{r}
Cities_temps_df %>%
  group_by(City) %>%
  summarise(Temp_mean = mean(Temperature),
            Temp_sd = sd(Temperature))
```
* Grouping the cities together under col(City)
* Calculating temperature mean and standard deviation from col(Temperature)
* Including the number of observations
* Calculating the standard error
* Defining a formula for standard error
```{r}
Cities_temps_df %>%
  group_by(City) %>%
  summarise(Temp_mean = mean(Temperature),
            Temp_sd = sd(Temperature),
            Temp_n = n()) %>%
  mutate(Temp_SE = Temp_sd/sqrt(Temp_n))
std_err <- function(x) {
  SE <- sd(x)/sqrt(length(x))
  return(SE)
}
```
* Grouping the cities together under col(City)
* Calculating temperature mean and standard error from col(Temperature)
```{r}
Cities_temps_df %>%
  group_by(City) %>%
  summarise(Temp_mean = mean(Temperature),
            Temp_SE = std_err(Temperature))
```
* Grouping the cities together under col(City)
* Calculating temperature mean and standard error from col(Temperature)
*Rounding the standard error to 3 digits
```{r}
Cities_temps_df %>%
  group_by(City) %>%
  summarise(Temp_mean = mean(Temperature),
            Temp_SE = round(std_err(Temperature), digits = 3))
```
* Grouping the cities together under col(City)
* Calculating temperature maximum and minimum values in each city
```{r}
Cities_temps_df %>%
  group_by(City) %>%
  summarise(Temp_min = min(Temperature),
            Temp_max = max(Temperature))
```
* Loading datasets to dataframes
* Craeting a new dataframe from 2 dataframes
* Joining the dataframes and kepping observations from df_moth_mass
* Looking at the dataset
* Grouping the Morph and SEX values together
* Calculating the mean of col(MASS)
```{r}
df_moth_mass <- read_csv("MOTH_MASS.csv")
df_moth_morph <- read_csv("MOTH_MORPHS.csv")

df_moth_joined <- left_join(df_moth_mass,
                            df_moth_morph)
df_moth_joined
df_moth_joined %>%
  group_by(SEX, MORPH) %>%
  summarise(Moth_mass_mean = mean(MASS))
```
* Loading a dataset to a dataframe
* Looking at the dataset
* Creating a dataframe
* Filtering out value 2014 from col(Year)
```{r}
df_weather <- read_csv("UKWeatherData.csv")

df_weather

df_weather
df_weather_new <- df_weather %>%
  filter(Year <= 2014)
```
* Creating a dataframe
* Arranging cols(Year, Month) in a descending order
* Looking at the new dataframe
* Keeping only col(Location)
```{r}
df_new_weather <- df_weather_new %>%
  arrange(desc(Year)) %>% 
  arrange(desc(Month))
df_new_weather

df_new_weather %>% distinct(Location)
```
* Creating a dataframe
* Keeping cols(Location, Year, Month, TempMax)
* Filtering col(Location) for Paisley
* Arranging col(TempMax) in a descending order
```{r}
Paisley_Max_Temp3 <- df_new_weather %>% 
  select(Location, Year, Month, TempMax) %>% 
  filter(Location == 'Paisley') %>% 
  arrange(desc(TempMax))
Paisley_Max_Temp3
```
* Creating a dataframe
* Grouping the years under col(Year)
* Calculating the mean of col(TempMax)
* Arranging col(TempMax) in a descending order
* Looking at the new dataframe
```{r}
Paisley_Max_Temp4 <- Paisley_Max_Temp3 %>%
  group_by(Year) %>%
  summarise(mean_TempMax = mean(TempMax)) %>% 
  arrange(desc(mean_TempMax))

Paisley_Max_Temp4
```
* Creating a dataframe
* Grouping the years under cols(Location, Year)
* Calculating the mean of col(TempMax)
* Arranging col(TempMax) in a descending order
* Looking at the new dataframe
```{r}
Paisley_Max_Temp6  <- df_new_weather %>%
  group_by(Location, Year) %>%
  summarise(mean_TempMax = mean(TempMax)) %>% 
  arrange(desc(mean_TempMax))

Paisley_Max_Temp6
```
* Creating a dataframe
* Filtering dataframe for year == 1999
* Grouping the months together under col(Month)
* Filtering out eventual Na-values in col(Rainfall)
* Calculating the mean and standard error of col(Rainfall)
* Rounding the standard error to 2 digits
* Looking at the dataframe
```{r}
RAINFALL_UK4  <- df_weather%>%
  filter(Year == 1999) %>% 
  group_by(Month) %>%
  filter(!is.na(Rainfall)) %>% 
  summarise(mean_rain = mean(Rainfall, na.rm = TRUE),
  se_rain = round(std_err(Rainfall), digits = 2))
RAINFALL_UK4
```
* Creating a dataframe
* grouping the location and year values
* Filtering out eventual Na-values in cols(TempMax,TempMin)
* Calculating temperature range
*Looking at dataframe
```{r}
Temprange_UK3  <- df_weather%>%
  group_by(Location, Year) %>%
  filter(!is.na(TempMax & TempMin)) %>% 
  summarise(Temprange = TempMax - TempMin)

Temprange_UK3 
```
* Creating a dataframe
* Filtering col(Year) for 200-2010
* Grouping the years un col(Year)
* Calculating the temperature range for each year and location
* Dropping eventual Na-values in col(Temprange)
* Ungrouping the grouping
* Looking at the dataframe
```{r}
HighTemp_UK2003 <- Temprange_UK3%>% 
  filter(Year >= 2000 & Year <= 2010)  %>% 
  group_by(Year) %>% 
  filter(Temprange == max(Temprange, na.rm = TRUE)) %>% 
  ungroup()
HighTemp_UK2003
```
* Creating col(HalYear) which shows which half of the year the observations were made on
```{r}
df_weather %>%
  mutate(HalfYear = ifelse(Month <= 6, "First", "Second"))
```
* Filtering col(Location) for value 'Stornoway'
* Filtering col(Year) for year range 2001-2010
* Creating col(HalfYear)
* Grouping the values in cols(Location, Year, HalfYear)
* Calculating the mean of monthly rainfall and dropping Na-values
```{r}
df_weather %>% 
  filter(Location == 'Stornoway') %>% 
  filter(Year >= 2001 & Year <= 2010) %>% 
    mutate(HalfYear = ifelse(Month <= 6, 'First', 'Second')) %>% 
    group_by(Location, Year, HalfYear) %>% 
    summarise(mean_rainfall = mean(Rainfall, na.rm = TRUE))
```
* Creating col(Quarter)
* Col(Quarter) shows on which quarter the observations were made
```{r}
df_weather %>%
  mutate(Quarter = ifelse(Month <= 3,1,
                          ifelse(Month <= 6,2,
                                 ifelse(Month <= 9,3,4))))
```
* Loading a dataset to a dataframe
* Looking at the dataset
* Joining 2 dataframes with col(Location)
* Keeping the obs. from df_weather
```{r}
df_stations <- read_csv("WeatherStations.csv")
df_stations
right_join(df_stations,df_weather,
           by = "Location")
```
* Filtering col(Year) for value 2014
* Selecting the cls(Location, Month, Rainfall)
* Grouping the locations together
* Calculating total rainfall 
* Dropping eventual Na-values
* Joining 2 dataframes
* Arranging col(total_rain) in a descending order
* Printing the entire dataframe
```{r}
df_weather %>%
  filter(Year == 2014) %>%
  select(Location, Month, Rainfall) %>%
  group_by(Location) %>%
  summarise(total_rain = sum(Rainfall, na.rm = TRUE)) %>%
  left_join(., df_stations) %>%
  arrange(desc(total_rain)) %>%
  print.data.frame()
```
* Link to Git_Hub repository
https://github.com/RoyBengtsson/Project_Vault.git 
